---
title: "R Solo Project - Habeeb"
output: html_notebook
---

```{r}
# Read in tidyverse library
library(tidyverse)
```

Question 1: Read in 'districts.csv' into a tibble named 'districts'. How many rows and columns does it contain?
```{r}
districts <- read_csv('../data/districts.csv')
```
Answer: The data set contains 146 rows and 27 columns.

Question 2: Notice that the first row corresponds to the whole State of Tennessee. Remove this row and save the result back to 'districts'.
```{r}
districts <- districts %>% 
  slice(-1)
```

Question 3: How many districts have a proficiency rate of at least 80% for both alg_1 and eng_1?
```{r}
districts %>%
  filter(alg_1 >= 80 & eng_1 >= 80)
```
Answer: 13 districts have a proficiency rate of at least 80% for both Algebra 1 and English 1.

Question 4: How many districts have a proficiency rate less than 50% for either alg_1 or eng_1?
```{r}
districts %>%
  filter(alg_1 < 50 | eng_1 < 50)
```
Answer: 8 districts have a proficiency rate less than 50% for either Algebra 1 or English 1.

Question 5: Which district has the lowest graduation rate?
```{r}
districts %>%
  select(system_name, grad) %>%
  arrange(grad)
```
Answer: Tennessee School for the blind has the lowest graduation rate at 11.1%.

Question 6: Which district has the highest dropout rate?
```{r}
districts %>%
  select(system_name, dropout) %>%
  arrange(desc(dropout))
```
Answer: Achievement School District has the highest dropout rate at 27.6%.

Question 7: Within the Mid Cumberland region, which district has the highest ACT composite?
```{r}
districts %>%
  filter(region == "Mid Cumberland") %>%
  select(system_name, act_composite) %>%
  arrange(desc(act_composite))
```
Answer: Williamson County has the highest ACT composite score within the Mid Cumberland region.

Question 8: Create a histogram showing the distribution of graduation rates. What can you say about this distribution?
```{r}
ggplot(districts, aes(x = grad)) +
  geom_histogram(bins = 25, fill = "blue", color = "black")
```
Answer: The data is skewed to the left and is uni modal around 85%. A majority of school districts see a graduation rate of 75% or greater.

Question 9: Create a scatter plot to compare alg_1 proficiency rates to alg_2 rates. What do you notice? Facet this plot by region. Does anything stand out when you facet the plots?
```{r}
ggplot(districts, aes(x = alg_1, y = alg_2)) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(districts, aes(x = alg_1, y = alg_2)) +
  geom_point() +
  facet_wrap(~region) +
  geom_smooth(method = "lm")
```
Answer: The scatter plot comparing the proficiency rates between Algebra 1 and Algebra 2 shows a positive correlation between the two variables. Once faceting the data by region, we see a neutral correlation in the Upper Cumberland region.

Question 10: Create a bar chart showing the total enrollment by region. Which region has the highest total enrollment? Which has the smallest?
```{r}
ggplot(districts, aes(x = region, y = enrollment)) +
  geom_col()
```

